---
title: "Preare gene expression data for TCGA SKCM samples"
output:
  html_document:
    df_print: paged
---


# Load required library
```{r warning=FALSE, echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
library(data.table)
```

# Read in raw data and gene length infomration
The raw data were downloaded from NCI GDC \url{https://portal.gdc.cancer.gov/}, using the HTSeq - Counts format. 

```{r}
datafolder = "TCGA_data/SKCM/gdc_download_20201012_225902.677830/"

# sample information
sample_info = read.table("TCGA_data/SKCM/gdc_sample_sheet.2020-10-12.tsv", 
                        header = TRUE, sep = "\t", as.is = TRUE)
dim(sample_info)
sample_info[1:3,]

table(sample_info$Data.Category)
table(sample_info$Data.Type)
table(sample_info$Project.ID)
table(table(sample_info$Case.ID))
table(table(sample_info$Sample.ID))
table(sample_info$Sample.Type)

tab.case = table(sample_info$Case.ID)
sam2 = sample_info[which(sample_info$Case.ID %in% names(tab.case)[tab.case > 1]),6:8]
sam2[order(sam2$Case.ID),]

# Info matrix on the genes: 60483 genes
geneinfo = fread(paste0(TCGAfolder, "gencode.v22.genes.txt"), header = TRUE, 
                 sep = "\t", drop = 8, na.strings = c("NA", ""))
head(geneinfo)

# Info matrix on the length of each gene: 60483 genes
load(paste0(TCGAfolder,"Gene_Lengths.RData"))
dim(GeneLengths.Mat)
head(GeneLengths.Mat)

table(GeneLengths.Mat$Gencode.ID == geneinfo$geneId)
```

# Step 1: Prepare count matrix
- read in each count data
- match the filename of count data to the sample ID
- check the gene ID of each new file is the same with the previous files
- combine the count data of all samples to a single count matrix
```{r}
# file paths for all individual count files
countfiles = paste0(datafolder, sample_info$File.ID, "/", sample_info$File.Name)
# save the gene names in the count files
gene.id = NA
n.genes = NA
# save the gene counts for each sample, col.names are sample ID
count.df = data.frame(matrix(numeric(), 0, length(countfiles),
                dimnames=list(c(), sample_info$Sample.ID)))

for (f in countfiles) {
  # read in the count file
  cf = fread(f, col.names = c("ensembl.ID", "counts"))
  # when reading in the first count file, save the gene names to gene.id, 
  # and save the counts to the corresponding column in the count matrix
  if (anyNA(gene.id)){
    gene.id = cf$ensembl.ID
    n.genes = nrow(cf)
    count.df[1:n.genes, match(f,countfiles)] = cf$counts
  } else {
    # starting from the second file, first check if number of rows match, 
    # then check whether its gene names match those in the first file
    # if so, save the counts to the corresponding column in the count matrix
    if(nrow(cf) != n.genes){ stop("number of genes do not match") }
    
    if (all(gene.id == cf$ensembl.ID)){
      count.df[,match(f,countfiles)] = cf$counts
    } else {
      # if not matching, print error message, skip this sample
      cat("wrong ensembl id in ", f, "\n")
      next
    }
  }
}

rownames(count.df) = gene.id
dim(count.df)
count.df[1:3,1:5]
anyNA(count.df)

rm(cf)

count.df[(nrow(count.df) - 5):nrow(count.df),1:5]

plot(unlist(count.df[60488,]), unlist(count.df[60485,]), 
     xlab="alignment_not_unique", ylab="ambiguous", pch=20, cex=0.6)

plot(unlist(count.df[60488,]), unlist(count.df[60484,]), 
     xlab="alignment_not_unique", ylab="no_feature", pch=20, cex=0.6)

plot(colSums(count.df[1:60483,]), unlist(count.df[60488,]),  
     xlab="total read-depth", ylab="alignment_not_unique", pch=20, cex=0.6)

names(count.df)[which(count.df[60488,] > 5e7)]

# Count.df has 60488 rows:
# - the last 5 rows are QC statistics, the others are counts for genes. 
# Match the rows of count.df to the rows of geneinfo by ensembl ID: 
# - 60483 gene ensembl ID in the data 
# - all matched to those in the gene info matrix in the same order
count.df = count.df[match(geneinfo$geneId, gene.id),]
dim(count.df)
count.df[c(1:3, nrow(count.df)), 1:5]
gene.id = row.names(count.df)

# save the raw matrix of count
saveRDS(count.df, file="TCGA_SKCM_raw_counts.rds")
```

# Step 2: Transform count data to TPM data
```{r}
calculate_TPM = function(count, gene_length) {
  if (nrow(count) != length(gene_length)) {
    stop("Number of rows of the count matrix does not match gene lengths!")
  }
  TPM = count / gene_length
  t(t(TPM)*1e6/colSums(TPM))
}
```

Genome-wide TPM normalization
```{r}
table(GeneLengths.Mat$Gencode.ID == rownames(count.df))

geneLength = GeneLengths.Mat$Exonic
SKCM_TPM   = calculate_TPM(count.df, geneLength) 
dim(SKCM_TPM)
SKCM_TPM[1:5,1:5]
```

# Step 3: Match gene ensembl ID to gene names
```{r}
# Some ensembl IDs miss hgnc symbols
missing.id = which(is.na(geneinfo$hgnc_symbol))
length(missing.id)
geneinfo.nomissing = geneinfo[-missing.id,]

# Some ensembl IDs have duplicated hgnc symbols
duplicated.id = which(duplicated(geneinfo.nomissing$hgnc_symbol))
length(duplicated.id)
geneinfo.filtered = geneinfo.nomissing[-duplicated.id,]
rm(geneinfo.nomissing)

dim(geneinfo.filtered)
geneinfo.filtered[1:2.]
table(geneinfo.filtered$geneId %in% gene.id)

table(gene.id == rownames(SKCM_TPM))

# Update SKCM_TPM to only genes with 1-1 mapping between 
# their ensembl ID and hgnc symbols, 35295 genes remained in the end. 
w2match = match(geneinfo.filtered$geneId, gene.id)
SKCM_TPM_filtered = data.frame(geneSymbol = geneinfo.filtered$hgnc_symbol) 
SKCM_TPM_filtered = cbind(SKCM_TPM_filtered, SKCM_TPM[w2match,])
dim(SKCM_TPM_filtered)
SKCM_TPM_filtered[1:5,1:5]

rm(SKCM_TPM)
```

Filter data and singature matrix using the genes shared between TCGA data and LM22 data: 535 genes in common. 
```{r}
# LM22 signature matrix
LM22_sigmat = read.table("../../_reference/LM22.txt", header = TRUE, 
                         sep = "\t", as.is = TRUE)
LM22_sigmat[1:5,1:5] # 547*23

genes2use = intersect(LM22_sigmat$Gene.symbol, SKCM_TPM_filtered$geneSymbol)
match.TCGA = match(genes2use, SKCM_TPM_filtered$geneSymbol)
match.LM22 = match(genes2use, LM22_sigmat$Gene.symbol)

SKCM_TPM_LM22 = SKCM_TPM_filtered[match.TCGA,]
dim(SKCM_TPM_LM22)
SKCM_TPM_LM22[1:2,1:5]

SKCM_LM22_sigmat = LM22_sigmat[match.LM22,]
dim(SKCM_LM22_sigmat)
SKCM_LM22_sigmat[1:2,1:5]
```

Filter data and singature matrix using the genes shared between TCGA data and the signature matrix `SF2018_sig_gene_matrix.txt`: 531 genes in common.

```{r}
SF2018_sigmat = read.table("../SF2018_analysis/output/SF2018_sig_gene_matrix.txt", 
                           header = TRUE, sep = "\t", as.is = TRUE)
dim(SF2018_sigmat)
SF2018_sigmat[1:5,1:5]

genes2use    = intersect(SF2018_sigmat$geneSymbol, SKCM_TPM_filtered$geneSymbol)
match.TCGA   = match(genes2use, SKCM_TPM_filtered$geneSymbol)
match.SF2018 = match(genes2use, SF2018_sigmat$geneSymbol)

SKCM_TPM_SF2018 = SKCM_TPM_filtered[match.TCGA,]
dim(SKCM_TPM_SF2018)
SKCM_TPM_SF2018[1:2,1:5]

SKCM_SF2018_sigmat = SF2018_sigmat[match.SF2018,] 
dim(SKCM_SF2018_sigmat)
SKCM_SF2018_sigmat[1:2,1:5]
```

# Generate CIBERSORT input

```{r}
CBSTfolder = "./CIBERSORT"
sig_file_LM22   = file.path(CBSTfolder, "signature_gene_LM22.txt")
mix_file_LM22   = file.path(CBSTfolder, "observed_TPM_LM22.txt")
sig_file_SF2018 = file.path(CBSTfolder, "signature_gene_SF2018.txt")
mix_file_SF2018 = file.path(CBSTfolder, "observed_TPM_SF2018.txt")

# generate matrices as input for CIBERSORT online
write.table(SKCM_TPM_LM22, file = mix_file_LM22, quote = FALSE,
            row.names = FALSE, sep="\t")

write.table(SKCM_LM22_sigmat, file = sig_file_LM22, quote = FALSE,
            row.names = FALSE, sep="\t")

write.table(SKCM_TPM_SF2018, file = mix_file_SF2018, 
            quote = FALSE, row.names = FALSE, sep="\t")

write.table(SKCM_SF2018_sigmat, file = sig_file_SF2018,
            quote = FALSE, row.names = FALSE, sep="\t")
```

# Run ICeDT

## Read in tumor purity information. 
```{r}
fnm = "pan_cancer_data/TCGA_mastercalls.abs_tables_JSedit.fixed.txt"
tumor_purity = fread(fnm)
dim(tumor_purity)
tumor_purity[1:2,]

length(unique(tumor_purity$array))
sample_id = names(SKCM_TPM_filtered)[-1]
sample_id[1:2]
sample_id = gsub(".", "-", sample_id, fixed = TRUE)
sample_id = substring(sample_id, 1, 15)
sample_id[1:2]

table(sample_id %in% tumor_purity$array)
sample2use  = intersect(sample_id, tumor_purity$array)
SKCM_purity = tumor_purity[match(sample2use, tumor_purity$array),]
dim(SKCM_purity)
SKCM_purity[1:2,]

summary(SKCM_purity$purity)
summary(SKCM_purity$ploidy)
summary(SKCM_purity$`Subclonal genome fraction`)

hist(SKCM_purity$purity, breaks=50, xlab="tumor purity", main="")
plot(SKCM_purity$purity, SKCM_purity$`Subclonal genome fraction`, 
     xlab="tumor purity", ylab="subclonal genome fraction")

table(SKCM_purity$purity < 0.95, useNA = "ifany")
table(SKCM_purity$purity < 0.99, useNA = "ifany")

SKCM_purity = SKCM_purity[which(SKCM_purity$purity < 0.99),]
dim(SKCM_purity)
```

## Run ICeDT using SF2018 signature matirx

```{r}
SKCM_TPM_SF2018_mtrx = data.matrix(SKCM_TPM_SF2018[,-1])
rownames(SKCM_TPM_SF2018_mtrx) = SKCM_TPM_SF2018$geneSymbol
SKCM_TPM_SF2018_mtrx = SKCM_TPM_SF2018_mtrx[,match(SKCM_purity$array, sample_id)]

SKCM_SF2018_sigmat_mtrx = data.matrix(SKCM_SF2018_sigmat[,-1])
rownames(SKCM_SF2018_sigmat_mtrx) = SKCM_SF2018_sigmat$geneSymbol

dim(SKCM_TPM_SF2018_mtrx)
SKCM_TPM_SF2018_mtrx[1:2,1:5]

dim(SKCM_SF2018_sigmat_mtrx)
SKCM_SF2018_sigmat_mtrx[1:2,1:5]

tcga_id = gsub(".", "-", colnames(SKCM_TPM_SF2018_mtrx), fixed = TRUE)
tcga_id = substring(tcga_id, 1, 15)
table(tcga_id == SKCM_purity$array)

date()
icedt_SF2018 = ICeDT(Y = SKCM_TPM_SF2018_mtrx, 
                     Z = SKCM_SF2018_sigmat_mtrx, 
                     tumorPurity = SKCM_purity$purity, refVar = NULL)
date()

summary(icedt_SF2018$cProp)

saveRDS(icedt_SF2018, "ICeDT/SKCM_ICeDT_SF2018.rds")
```

## Run ICeDT using LM22 signature matirx

```{r}
SKCM_TPM_LM22_mtrx = data.matrix(SKCM_TPM_LM22[,-1])
rownames(SKCM_TPM_LM22_mtrx) = SKCM_TPM_LM22$geneSymbol
SKCM_TPM_LM22_mtrx = SKCM_TPM_LM22_mtrx[,match(SKCM_purity$array, sample_id)]

SKCM_LM22_sigmat_mtrx = data.matrix(SKCM_LM22_sigmat[,-1])
rownames(SKCM_LM22_sigmat_mtrx) = SKCM_LM22_sigmat$Gene.symbol

dim(SKCM_TPM_LM22_mtrx)
SKCM_TPM_LM22_mtrx[1:2,1:5]

dim(SKCM_LM22_sigmat_mtrx)
SKCM_LM22_sigmat_mtrx[1:2,1:5]

tcga_id = gsub(".", "-", colnames(SKCM_TPM_LM22_mtrx), fixed = TRUE)
tcga_id = substring(tcga_id, 1, 15)
table(tcga_id == SKCM_purity$array)

date()
icedt_LM22 = ICeDT(Y = SKCM_TPM_LM22_mtrx, 
                   Z = SKCM_LM22_sigmat_mtrx, 
                   tumorPurity = SKCM_purity$purity, refVar = NULL)
date()

summary(icedt_LM22$cProp)

saveRDS(icedt_LM22, "ICeDT/SKCM_ICeDT_LM22.rds")
```


```{r}
gc()
sessionInfo()
```
