---
title: "Compare cell type proportion estimates for TCGA SKCM samples"
output:
  html_document:
    df_print: paged
    keep_md: TRUE
---


## Load required library
```{r warning=FALSE, echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
library(data.table)
library(stringr)
library(MASS)

library(ggplot2)
library(gridExtra)
library(cowplot)
library(ggcorrplot)
library(GGally)
library(ggpubr)
theme_set(theme_bw())

library(tidyr)
```

## read in cell type proportion estimates

Read in cell type proportion estimates by EMeth using methylaiton data and the estimates by CIBERSORT, using two references: one was LM22, the default signature of CIBERSORT, and the other one was 11 cell types generated using scRNA-seq ata from SF2018


```{r}
load("TCGA_results/deconv_methy_SKCM.RData")
ls()
rho_methy = rho

load("TCGA_results/deconv_expr_SKCM.RData")
ls()
rho_LM22 = deconv_expr

dim(rho_methy)
rho_methy[1,,]

dim(rho_LM22)
rho_LM22[1:2,]

table(rownames(rho_LM22) == dimnames(rho_methy)[[1]])

fnm = "CIBERSORTx_SKCM_SF2018_signature_Adjusted.txt"
rho_SF11 = as.data.frame(fread(paste0("_cibersortx_results/", fnm)))
dim(rho_SF11)
rho_SF11[1:2,]

rho_SF11$patient_id = substr(rho_SF11$Mixture, 9, 12)
```

## read in tumor purity information
```{r}
dir0 = "TCGA_results/clinical_data/"
fnm  = paste0(dir0, "TCGA_mastercalls.abs_tables_JSedit.fixed.txt")
tcga_purity = fread(fnm)

dim(tcga_purity)
tcga_purity[1:2,]

rho_SF11$array = substr(rho_SF11$Mixture, 1, 15)
rho_SF11$array = gsub(".", "-", rho_SF11$array, fixed=TRUE)
dim(rho_SF11)
rho_SF11[1:2,(ncol(rho_SF11) -5):ncol(rho_SF11)]

table(rho_SF11$array %in% tcga_purity$array)

array2use   = intersect(rho_SF11$array, tcga_purity$array)
rho_SF11    = rho_SF11[match(array2use, rho_SF11$array),]
tcga_purity = tcga_purity[match(array2use, tcga_purity$array),]

dim(rho_SF11)
dim(tcga_purity)
table(rho_SF11$array == tcga_purity$array)
rho_SF11$purity = tcga_purity$purity

patient_id = intersect(rownames(rho_LM22), rho_SF11$patient_id)
length(patient_id)

# note that if a patient has more than one sample, we just 
# use the first one
rho_SF11  = rho_SF11[match(patient_id,  rho_SF11$patient_id),]
rho_LM22  = rho_LM22[match(patient_id,  rownames(rho_LM22)),]
rho_methy = rho_methy[match(patient_id, dimnames(rho_methy)[[1]]),,]

dim(rho_SF11)
dim(rho_LM22)
dim(rho_methy)
```


## collapse cell types proportion estimates to fewer cell types

```{r}
ct_list = list()
ct_list[["B"]] = c("B_cells", "Plasma_cells")
ct_list[["CD4T"]] = c("CD4T_memory")
ct_list[["CD8T"]] = c("CD8T_B", "CD8T_G")
ct_list[["Treg"]] = c("Tregs")
ct_list[["NK"]]   = c("NK1", "NK2", "NK3")
ct_list[["Monocyte"]]   = c("Monocytes_Macrophages", "Dendritic_cells")
cell_size_factors = c(0.4, 0.4, 0.4, 0.4, 0.42, 1.40)
names(cell_size_factors) = c("B", "CD4T", "CD8T", "Treg", "NK", "Monocyte")

rho_expr_sf11 = matrix(NA, nrow = nrow(rho_SF11), ncol = length(ct_list))
colnames(rho_expr_sf11) = names(ct_list) 
rownames(rho_expr_sf11) = rho_SF11$Mixture

for(ct1 in names(ct_list)){
  cts = ct_list[[ct1]]
  rho_expr_sf11[,ct1] = rowSums(as.matrix(rho_SF11[,cts]))
}

rho_expr_sf11 = rho_expr_sf11 / rowSums(rho_expr_sf11)

dim(rho_expr_sf11)
rho_expr_sf11[1:2,]
```

## correct for tumor purity
```{r}
eta = rho_SF11$purity
summary(eta)
eta[which(eta > 0.99)] = 0.99

rho_expr_sf11 = diag(1-eta) %*% rho_expr_sf11
rownames(rho_expr_sf11) = rho_SF11$patient_id
dim(rho_expr_sf11)
rho_expr_sf11[1:2,]

rho_SF11 = rho_expr_sf11
```


## compare cell type proportion between methods
```{r corrplot, fig.height=3, fig.width=3.6}
rho_SF11  = rho_SF11[,order(colnames(rho_SF11))]
rho_LM22  = rho_LM22[,order(colnames(rho_LM22))]
rho_EMeth = rho_methy[,,"EMeth"]
rho_EMeth = rho_EMeth[,order(colnames(rho_EMeth))]

ggcorrplot(cor(rho_SF11, rho_LM22), tl.cex = 10) + 
  labs(x = 'SF11', y = 'LM22') +
  theme(
    axis.title.x = element_text(size=12, face="bold"),
    axis.title.y = element_text(angle = 90, size=12, face="bold")
  )

ggcorrplot(cor(rho_SF11, rho_EMeth), tl.cex = 10) + 
  labs(x = 'SF11', y = 'EMeth') + 
  theme(
    axis.title.x = element_text(size=12, face="bold"),
    axis.title.y = element_text(angle = 90, size=12, face="bold")
  )

ggcorrplot(cor(rho_LM22, rho_EMeth), tl.cex = 10) + 
  labs(x = 'LM22', y = 'EMeth') + 
  theme(
    axis.title.x = element_text(size=12, face="bold"),
    axis.title.y = element_text(angle = 90, size=12, face="bold")
  )
```


## compare cell type proportion between methylation and LM22
```{r}
methods = c("EMeth","svr","ls","rls","qp")
cellTypes = dimnames(rho_methy)[[2]]

utypes = intersect(cellTypes, colnames(rho_LM22))
utypes

cor_LM22 <- matrix(NA,nrow = length(utypes), ncol = length(methods))
colnames(cor_LM22) <- methods
rownames(cor_LM22) <- utypes

err_LM22 = cor_LM22

for(i in 1:length(utypes)){
  rho_e_i = rho_LM22[,utypes[i]]
  
  cor_LM22[i,] <- sapply(1:length(methods), FUN = function(j){
    cor(rho_methy[,utypes[i],methods[j]], rho_e_i, use="pair")
  })
  err_LM22[i,] <- sapply(1:length(methods), FUN = function(j){
    sqrt(mean((rho_methy[,utypes[i],methods[j]] - rho_e_i)^2, na.rm=TRUE))
  }) 
}
```

## compare cell type proportion between methylation and SF11

```{r}
utypes = intersect(cellTypes, colnames(rho_SF11))
utypes

cor_SF11 <- matrix(NA,nrow = length(utypes), ncol = length(methods))
colnames(cor_SF11) <- methods
rownames(cor_SF11) <- utypes

err_SF11 = cor_SF11

for(i in 1:length(utypes)){
  rho_e_i = rho_SF11[,utypes[i]]
  
  cor_SF11[i,] <- sapply(1:length(methods), FUN = function(j){
    cor(rho_methy[,utypes[i],methods[j]], rho_e_i, use="pair")
  })
  err_SF11[i,] <- sapply(1:length(methods), FUN = function(j){
    sqrt(mean((rho_methy[,utypes[i],methods[j]] - rho_e_i)^2, na.rm=TRUE))
  }) 
}
```

## Check the comparison results

```{r barplot, fig.height=3, fig.width=6}

cor_LM22
err_LM22

cor_SF11
err_SF11

df_cor_LM22 = pivot_longer(data.frame(cor_LM22, cell_type=rownames(cor_LM22)), 
                           !cell_type, names_to = "method", values_to = "corr")

p_cor_LM22 <- ggplot(data=df_cor_LM22, aes(x=cell_type, y=corr, fill=method)) +
  geom_bar(stat="identity", color="black", position=position_dodge()) + 
  theme_minimal()

df_cor_SF11 = pivot_longer(data.frame(cor_SF11, cell_type=rownames(cor_SF11)), 
                           !cell_type, names_to = "method", values_to = "corr")

p_cor_SF11 <- ggplot(data=df_cor_SF11, aes(x=cell_type, y=corr, fill=method)) +
  geom_bar(stat="identity", color="black", position=position_dodge()) + 
  theme_minimal()

p_cor_LM22
p_cor_SF11
```


## Check a few cell types
```{r scatters, fig.height=5, fig.width=5}
Treg_e_LM22 = rho_LM22[,"Treg"]
Treg_e_SF11 = rho_SF11[,"Treg"]
Treg_methy  = rho_methy[, "Treg", "EMeth"]

df1 = data.frame(Treg_e_LM22, Treg_e_SF11, Treg_methy)
ggpairs(df1)

CD8T_e_LM22 = rho_LM22[,"CD8T"]
CD8T_e_SF11 = rho_SF11[,"CD8T"]
CD8T_methy  = rho_methy[, "CD8T", "EMeth"]

df1 = data.frame(CD8T_e_LM22, CD8T_e_SF11, CD8T_methy)
ggpairs(df1)
```

```{r}
gc()
sessionInfo()
```


